---
title: "Canopy Height Models"
---

```{r, echo = FALSE}
library(rgl)

r3dDefaults <- rgl::r3dDefaults
m <- structure(c(0.921, -0.146, 0.362, 0, 0.386, 0.482, -0.787, 0, 
                -0.06, 0.864, 0.5, 0, 0, 0, 0, 1), .Dim = c(4L, 4L))
r3dDefaults$FOV <- 50
r3dDefaults$userMatrix <- m
r3dDefaults$zoom <- 0.75

knitr::opts_chunk$set(
  comment =  "#>", 
  collapse = TRUE,
  fig.align = "center")

rgl::setupKnitr(autoprint = TRUE)
```

## Relevant Resources

[Code](https://github.com/tgoodbody/lidRtutorial/blob/main/code/tutorials/04_chm.R)

[lidRbook section](https://r-lidar.github.io/lidRbook/chm.html)

## Overview

This code demonstrates the creation of a Canopy Height Model (CHM) using LiDAR data. It shows different algorithms for generating CHMs and provides options for adjusting resolution, sub-circle size, and filling empty pixels.

## Environment

```{r clear_warnings, warnings = FALSE, message = FALSE}
# Clear environment
rm(list = ls(globalenv()))

# Load packages
library(lidR)
library(microbenchmark)
library(terra)
```

## Data Preprocessing

In this section, we load the LiDAR data, set a random fraction filter to reduce point density, and visualize the resulting LiDAR point cloud.

```{r load_lidar_data}
# Load LiDAR data and reduce point density
las <- readLAS(files = "data/MixedEucaNat_normalized.laz", filter = "-keep_random_fraction 0.4 -set_withheld_flag 0")

# Visualize the LiDAR point cloud
plot(las)
```

## Point-to-Raster Based Algorithm

In this section, we demonstrate a simple method for generating Canopy Height Models (CHMs) that assigns the elevation of the highest point to each pixel.

```{r point_to_raster_algorithm}
# Generate the CHM using a simple point-to-raster based algorithm
chm <- grid_canopy(las = las, res = 2, algorithm = p2r())

# Visualize the CHM
plot(chm, col = height.colors(50))
```

In the first code chunk, we generate a Canopy Height Model (CHM) using a point-to-raster based algorithm. The `grid_canopy` function with the `p2r()` algorithm assigns the elevation of the highest point within each grid cell to the corresponding pixel. The resulting CHM is then visualized using the `plot()` function.

```{r point_to_raster_algorithm_equivalent}
# The above method is strictly equivalent to using pixel_metrics to compute max height
chm <- pixel_metrics(las = las, func = ~max(Z), res = 2)

# Visualize the CHM
plot(chm, col = height.colors(50))
```

The code chunk above shows that the point-to-raster based algorithm is equivalent to using `pixel_metrics` with a function that computes the maximum height (`max(Z)`) within each grid cell. The resulting CHM is visualized using the `plot()` function.

```{r point_to_raster_algorithm_optimized}
# However, the grid_canopy algorithm is optimized
microbenchmark::microbenchmark(canopy = grid_canopy(las = las, res = 1, algorithm = p2r()),
                               metrics = pixel_metrics(las = las, func = ~max(Z), res = 1),
                               times = 10)
```

The above code chunk uses `microbenchmark::microbenchmark()` to compare the performance of the `grid_canopy()` function with `p2r()` algorithm and `pixel_metrics()` function with `max(Z)` for maximum height computation. It demonstrates that the `grid_canopy()` function is optimized for generating CHMs.

```{r point_to_raster_algorithm_higher_resolution}
# Increasing the resolution results in fewer empty pixels
chm <- grid_canopy(las = las, res = 1, algorithm = p2r())
plot(chm, col = height.colors(50))
```

By increasing the resolution of the CHM (reducing the grid cell size), we get a more detailed representation of the canopy with fewer empty pixels.

```{r point_to_raster_algorithm_subcircle}
# Using the 'subcircle' option turns each point into a disc of 8 points with a radius r
chm <- grid_canopy(las = las, res = 0.5, algorithm = p2r(subcircle = 0.15))
plot(chm, col = height.colors(50))
```

The `grid_canopy()` function with the `p2r()` algorithm allows the use of the `subcircle` option, which turns each LiDAR point into a disc of 8 points with a specified radius. This can help to capture more fine-grained canopy details in the resulting CHM.

```{r point_to_raster_algorithm_larger_subcircle_radius}
# Increasing the subcircle radius, but it may not have meaningful results
chm <- grid_canopy(las = las, res = 0.5, algorithm = p2r(subcircle = 0.8))
plot(chm, col = height.colors(50))
```

Increasing the `subcircle` radius may not necessarily result in meaningful CHMs, as it could lead to over-smoothing or loss of important canopy information.

```{r point_to_raster_algorithm_fill_empty_pixels}
# We can fill empty pixels using TIN interpolation
chm <- grid_canopy(las = las, res = 0.5, algorithm = p2r(subcircle = 0.15, na.fill = tin()))
plot(chm, col = height.colors(50))
```

The `p2r()` algorithm also allows filling empty pixels using TIN (Triangulated Irregular Network) interpolation, which can help in areas with sparse LiDAR points to obtain a smoother CHM.

## Triangulation Based Algorithm

In this section, we demonstrate a triangulation-based algorithm for generating CHMs.

```{r triangulation_algorithm}
# Triangulation of first returns to generate the CHM
chm <- grid_canopy(las = las, res = 1, algorithm = dsmtin())
plot(chm, col = height.colors(50))
```

The `grid_canopy()` function with the `dsmtin()` algorithm generates a CHM by performing triangulation on the first returns from the LiDAR data. The resulting CHM represents the surface of the canopy.

```{r triangulation_algorithm_higher_resolution}
# Increasing the resolution results in a more detailed CHM
chm <- grid_canopy(las = las, res = 0.5, algorithm = dsmtin())
plot(chm, col = height.colors(50))
```

Increasing the resolution of the CHM using the `res` argument provides a more detailed representation of the canopy, capturing finer variations in the vegetation.

```{r triangulation_algorithm_pitfree}
# Using the Khosravipour et al. pit-free algorithm with specified thresholds and maximum edge length
thresholds <- c(0, 5, 10, 20, 25, 30)
max_edge <- c(0, 1.35)
chm <- grid_canopy(las = las, res = 0.5, algorithm = pitfree(thresholds, max_edge))
plot(chm, col = height.colors(50))
```

The `grid_canopy` function can also use the Khosravipour et al. pit-free algorithm with specified height thresholds and a maximum edge length to generate a CHM. This algorithm aims to correct depressions in the CHM surface.

```{r triangulation_algorithm_subcircle}
# Using the 'subcircle' option with the pit-free algorithm
chm <- grid_canopy(las = las, res = 0.5, algorithm = pitfree(thresholds, max_edge, 0.1))
plot(chm, col = height.colors(50))
```

The `subcircle` option can be used with the pit-free algorithm to create finer CHMs with subcircles for each LiDAR point, similar to the point-to-raster based algorithm.

## Post-Processing

Usually, CHMs can be post-processed by smoothing or other manipulations. Here, we demonstrate post-processing using the `terra` package and the `focal()` function for smoothing.

```{r post_processing}
# Post-process the CHM using the 'terra' package and focal() function for smoothing
ker <- matrix(1, 3, 3)
schm <- terra::focal(chm, w = ker, fun = mean, na.rm = TRUE)

# Visualize the smoothed CHM
plot(schm, col = height.colors(50))
```

Post-processing of CHMs often involves smoothing to improve data quality or remove noise. In this example, we use the `terra::focal()` function to perform a focal mean smoothing on the CHM, resulting in a smoother representation of the canopy height.

## Conclusion

This tutorial covered different algorithms for generating Canopy Height Models (CHMs) from LiDAR data using the `lidR` package in R. It includes point-to-raster-based algorithms and triangulation-based algorithms, as well as post-processing using the `terra` package. The code chunks are well-labeled to help the audience navigate through the tutorial easily.
